---
title: "How to carry column metadata in pivot_longer"
author: "Nick Carruthers"
date: "15/06/2020"
output: md_document
# knit: (function(input_file, encoding) {
#   rmarkdown::render(input_file,
#  encoding=encoding,
#  output_file=file.path(dirname(input_file), 'index.html'))})
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Background:
Pivoting data can be a pain point in bioinformatics workflows.  Lots of bioinformatics software are tied to the **wide** format with data spread out among multiple columns while the whole tidyverse/ggplot system requires **long** data with as few columns as possible.  Becoming proficient at switching your data to **long** format has several benefits.  (1) It provides a unified format for any required data manipulations and summarizations making them faster to write and easier to read and (2) it is the required input format for the ggplot system.  In R [the tidyverse](https://www.tidyverse.org/) provides the tools to interchange **wide** and **long** data.    

#### The Problem:  
Typically subjects in bioinformatics datasets (columns) will have associated metadata like treatments and indicators of groups or replicates.  Any metadata that corresponds to rows can be easily added to the data.frame to be pivoted (eg. with `cbind`).  But if there are column metadata they have to be added manually after the pivot.  

#### The Solution:
There are a couple of ways to do this.  The way I've settled on is to have a table of target meta-data and use a join after the pivot to connect it to the data.  I find myself doing this repeatedly in almost all of my analyses but it's a solution I stumbled on by trial and error.  I have never seen it spelled out explicitly anywhere so here it is.  

```{r include = FALSE}
library(tidyverse)

library(devtools)
install_github(c("duncantl/XMLRPC", "duncantl/RWordPress"))
```

```{r}
head(relig_income)
```

#### First, create the metadata.  
I'll use the `relig_income` dataset as an example.  I will demonstrate how to add both row metadata (easy) and column metadata (bit tricky).  For row metadata I will add a new column for religion class that will be defined randomly and for column metadata I will group income levels into low, medium, high and unknown listed in a separate `data.frame`.  Note that this method relies on linking data column names to metadata so check the metadata table carefully!  

To add the row metadata I simply add a new column to the `relig_income` table with my random values.  For the column metadata I will make a new `data.frame`.  

```{r}
## Row metadata
set.seed(10)
relig_income$religionClass <- 
  sample(c("A", "B", "C"), nrow(relig_income), replace = TRUE)

## Column metadata
columnMetadata <- data.frame(
  income = c(colnames(relig_income)[
    grepl("0", colnames(relig_income))],
    "Don't know/refused"),
  incomeGroup = c(rep("low", 3), rep("medium", 3), 
            rep("high", 3), "Don't know/refused"))
columnMetadata
```

#### Step 1: pivot_longer as usual  

Don't forget to exclude the new `religionClass` column from the pivot.  

```{r}
relig_income %>%
  pivot_longer(-c(religion, religionClass), names_to = "income", values_to = "count")
```

#### Step 2:  join the column metadata  
All metadata columns will be added automatically with this step.  

```{r}
relig_income %>%
  pivot_longer(-c(religion, religionClass), names_to = "income", values_to = "count") %>%
  inner_join(columnMetadata, by = "income")

```

#### Step 3 (optional):  Convert character data to ordered factors to control plotting order

```{r}
relig_income %>%
  pivot_longer(-c(religion, religionClass), names_to = "income", values_to = "count") %>%
  inner_join(columnMetadata, by = "income") %>%
  mutate(income = ordered(income, levels = columnMetadata$income))
```


Finally look at the mapping to ensure it worked.  Unfortunately table doesn't play well with the `%>%` operator so this step is a bit inelegant.  

```{r}
test <- relig_income %>%
  pivot_longer(-c(religion, religionClass), names_to = "income", values_to = "count") %>%
  inner_join(columnMetadata, by = "income") %>%
  mutate(income = ordered(income, levels = columnMetadata$income))

table(test$income, test$incomeGroup)
```

#### The metadata columns are now available
We can plot the data summarized by our arbitrary grouping of religions and colored by our grouped income levels.  Order the income classes to make a sensible presentation

```{r fig.width = 8, fig.height = 8}
relig_income %>%
  pivot_longer(-c(religion, religionClass), names_to = "income", values_to = "count") %>%
  inner_join(columnMetadata, by = "income") %>%
  mutate(income = ordered(income, levels = columnMetadata$income)) %>%
  mutate(incomeGroup = ordered(incomeGroup, levels = c("low", "medium", "high", "Don't know/refused"))) %>%
  group_by(religionClass, income, incomeGroup) %>%
  summarize(meanCount = mean(count), .groups =  "drop_last") %>%
  ggplot(aes(x = income, y = meanCount, fill = incomeGroup)) +
  geom_col() +
  facet_wrap(vars(religionClass)) + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

```

